\documentclass{article} %


\usepackage{etoolbox}
\usepackage{comment}
\newtoggle{icml}
\togglefalse{icml}

\newcommand{\icml}[1]{\iftoggle{icml}{#1}{}}
\newcommand{\arxiv}[1]{\iftoggle{icml}{}{#1}}
\newcommand{\loose}{\looseness=-1}

\icml{
  \usepackage{icml2026}
}

\input{arxiv_style}
\input{macros}
\input{prev_macros.tex}
\let\underbar\undefined
\input{widebar}

\usepackage[utf8]{inputenc} %
\usepackage[T1]{fontenc}    %
\usepackage{url}            %
\usepackage{booktabs}       %
\usepackage{amsfonts}       %
\usepackage{nicefrac}       %
\usepackage{microtype}      %
\usepackage{makecell}
\usepackage{enumitem}
\usepackage{breakcites}
\usepackage{mathrsfs}

\usepackage[normalem]{ulem}

\arxiv{
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{verbatim}
}

\icml{
\newcommand{\State}{\STATE}
\newcommand{\For}{\FOR}
\newcommand{\EndFor}{\ENDFOR}
}

\usepackage{multicol}
\usepackage{colortbl}
\usepackage{setspace}
\usepackage{transparent}
\usepackage{upgreek}

\usepackage{inconsolata}
\usepackage[scaled=.90]{helvet}
\usepackage{xspace}

\icml{
\setlist[enumerate]{leftmargin=*}
\setlist[itemize]{leftmargin=*}
}

\usepackage{graphicx}
\icml{\usepackage{subfigure}}

\usepackage[suppress]{color-edits}
\addauthor{fc}{blue}
\addauthor{sc}{orange}
\addauthor{sr}{red}



\let\oldparagraph\paragraph
\newcommand{\paragraphi}[1]{\oldparagraph{\emph{#1}.}}
\renewcommand{\paragraph}[1]{\oldparagraph{#1.}}

\newcommand{\textbfc}[1]{\textbf{\textcolor{blue!40!black}{#1}}}


\makeatletter
\g@addto@macro\appendix{%
  \crefalias{section}{appendixsection}%
  \crefalias{subsection}{appendixsubsection}%
  \crefalias{subsubsection}{appendixsubsubsection}%
}
\makeatother
\crefname{appendixsection}{Appendix}{Appendices}
\Crefname{appendixsection}{Appendix}{Appendices}
\crefname{appendixsubsection}{Appendix}{Appendices}
\Crefname{appendixsubsection}{Appendix}{Appendices}
\crefname{appendixsubsubsection}{Appendix}{Appendices}
\Crefname{appendixsubsubsection}{Appendix}{Appendices}


\begin{document}

\icml{
\twocolumn[
  \icmltitle{High-accuracy sampling for diffusion models and log-concave distributions}



  \icmlsetsymbol{equal}{*}

  \begin{icmlauthorlist}
    \icmlauthor{Firstname1 Lastname1}{equal,yyy}
    \icmlauthor{Firstname2 Lastname2}{equal,yyy,comp}
    \icmlauthor{Firstname3 Lastname3}{comp}
    \icmlauthor{Firstname4 Lastname4}{sch}
    \icmlauthor{Firstname5 Lastname5}{yyy}
    \icmlauthor{Firstname6 Lastname6}{sch,yyy,comp}
    \icmlauthor{Firstname7 Lastname7}{comp}
    \icmlauthor{Firstname8 Lastname8}{sch}
    \icmlauthor{Firstname8 Lastname8}{yyy,comp}
  \end{icmlauthorlist}

  \icmlaffiliation{yyy}{Department of XXX, University of YYY, Location, Country}
  \icmlaffiliation{comp}{Company Name, Location, Country}
  \icmlaffiliation{sch}{School of ZZZ, Institute of WWW, Location, Country}

  \icmlcorrespondingauthor{Firstname1 Lastname1}{first1.last1@xxx.edu}
  \icmlcorrespondingauthor{Firstname2 Lastname2}{first2.last2@www.uk}

  \icmlkeywords{Machine Learning, ICML}

  \vskip 0.3in
]



\printAffiliationsAndNotice{}  %
}

\arxiv{
  \title{High-accuracy sampling for diffusion models and log-concave distributions}
\author{
  Fan Chen \\ {\small MIT} \\ {\small \texttt{fanchen@mit.edu}} 
  \and Sinho Chewi \\ {\small Yale University} \\ {\small \texttt{sinho.chewi@yale.edu}} 
  \and Constantinos Daskalakis \\ {\small MIT} \\ {\small \texttt{costis@csail.mit.edu}}
  \and Alexander Rakhlin \\ {\small MIT} \\ {\small \texttt{rakhlin@mit.edu}}
}
  \maketitle
}

\begin{abstract}
We present algorithms for diffusion model sampling which obtain $\delta$-error in $\mathrm{polylog}(1/\delta)$ steps, given access to $\widetilde O(\delta)$-accurate score estimates in $L^2$. This is an exponential improvement over all previous results.
Specifically, under minimal data assumptions, the complexity is $\widetilde O(d_\star\,\mathrm{polylog}(1/\delta))$ where $d_\star$ is the intrinsic dimension of the data. Further, under a non-uniform $L$-Lipschitz condition, the complexity reduces to $\widetilde O(L\,\mathrm{polylog}(1/\delta))$. 
Our approach also yields the first $\mathrm{polylog}(1/\delta)$ complexity sampler for general log-concave distributions using only gradient evaluations.
\end{abstract}


\section{Introduction}


What is the complexity of sampling from a continuous probability distribution, given access to evaluations of the gradient of the log-density?
In particular, can one design algorithms whose iteration complexity scales as $\polylog(1/\delta)$, where $\delta$ is the target accuracy, or must they necessarily take $\poly(1/\delta)$ steps?

Complexity bounds which scale as $\polylog(1/\delta)$, indicating that algorithms converge exponentially fast, are known as ``high-accuracy'' guarantees, as they ensure that one can draw an extremely accurate sample without too many iterations.
When access to the log-density of the target distribution is available, and not just its gradient, then high-accuracy samplers abound, based on accept-reject mechanisms such as rejection sampling or the Metropolis--Hastings filter.

Without density evaluations, the answer is less clear.
In this setting, existing sampling methods are typically based on discretizations of stochastic differential equations, and the need to control the discretization error precludes high-accuracy guarantees.
(Note that this is unlike the setting of optimization, in which gradient descent enjoys a high-accuracy guarantee under strong convexity---there, discretization does not bias the algorithm.)
A notable exception is discretization of piecewise deterministic Markov processes (PDMPs), for which~\citet{LuWan22Zigzag} show that $\polylog(1/\delta)$ evaluations of the gradient suffice~\fcedit{from a warm start}.

This question has become particularly interesting in light of recent developments on diffusion-based generative modeling.
Such models are based on implementing a certain reverse Markov process, in which each iteration requires evaluation of a score function---the gradient of the log-density along a diffusion process.
Since these models only learn the score function and not the density function itself (unlike earlier approaches such as energy-based models), it is natural to ask what the best achievable complexity is using score evaluations alone.

More precisely, we are interested in bounding the number of steps (and queries to the score function) required to sample from a distribution $\phat$ such that
\begin{align}\label{eq:goal}
  D(\pdata,\phat)\leq \delta+ \Capx\cdot  \epssc\,,
\end{align}
where $D$ is a measure of discrepancy, $\delta\in(0,1)$ is the target accuracy, $\epssc$ measures the error in the score function estimates, and $\Capx$ is a target approximation factor.


The initial work of~\citet{Chen+23SGM, LeeLuTan23SGMGeneral} showed that with $L^2$-accurate score estimates, Denoising Diffusion Probabilistic Models (DDPMs) achieve a query complexity of $1/\delta^2$ in total variation distance for sampling from an early stopped distribution, provided that $\pdata$ has bounded support, with an overhead that is polynomial in the dimension and radius, and with $\Capx = \wt O(1)$.
By a standard argument, this can be converted to guarantees for sampling from $\pdata$ itself in a weaker metric.
Importantly, these results impose almost no assumptions on the data distribution, providing strong theoretical justification for diffusion models. 


Since then, there has been an explosion of works aimed at extending and refining these guarantees~\citep[e.g.,][]{chen2023improved,li2023towards,Ben+24Diffusion,gao2025wasserstein,ConDurSil25KLDiffusion, LiYan25Diffusion,Chen+23FlowODE,li2024sharp, GaoZhu25PFODE, huang2025convergence, JaiZha26SharpKL}, which we cannot fully survey here.
However, it is known that a complexity of $\Omega(1/\delta)$ is unimprovable for DDPM~\citep{jiao2025optimal}, which motivates changing the algorithm.
Under minimal assumptions, the algorithm of~\citet{li2024provable} achieves a query complexity of $1/\delta^{1/2}$.
Other works achieved similar speed-ups assuming that the Jacobian of the score error is bounded~\citep{li2024accelerating,li2025faster}.

A recent line of works~\citep[e.g.,][]{huang2025convergence, huang2025fast, li2025faster} studied higher-order discretization methods, with the resulting query complexity scaling roughly as $\CP\, d^{1+1/p}/\delta^{1/p}$,
where $p\geq 1$ is the chosen acceleration order, and $\CP$ hides dependence on other problem parameters\footnote{For example, in the results of \citet{huang2024reverse,huang2025convergence,huang2025fast}, $\CP$ hides polynomial dependence on higher-order Lipschitz parameters or the diameter of the support of $\pdata$.} 
along with an implicit (often exponential) dependence on $p$. These results show that diffusion sampling can be performed with sub-polynomial dependence on $1/\delta$,\footnote{
  Roughly, we should expect $\CP\geq e^{\Omega(p)}$. Hence, the complexity is at least $d\exp(O(\sqrt{\log(d/\delta)}))$ with the best choice of acceleration order $p$, which is better than $\poly(1/\delta)$ but much worse than $\poly\log(1/\delta)$.
}
but they still fall well short of the desired poly-logarithmic complexity.\footnote{Furthermore, the approximation factor $\Capx$ in \citet{huang2025convergence,huang2025fast} scales polynomially in the early stopping time, and \citet{li2025faster} additionally relies on the second-order score error $\epsjac$, making the results sensitive to score errors.}

Meanwhile, \citet{huang2024reverse, wainwright2025score} propose to use density evaluations to achieve high-accuracy samplers, and~\citet{Hua+25QuantizedDiffusion} propose a high-accuracy sampler which requires learning a certain ``quantized'' score.
As discussed above, these are not compatible with the current practice of diffusion sampling, which only learns score estimates.

Therefore, we state our main question of interest:
\begin{center}
    \emph{Is there a diffusion model sampler which achieves a high-accuracy guarantee using only score evaluations, under minimal assumptions on both the data distribution and the score error?}
\end{center}





\subsection{Our contribution}


We answer this question affirmatively via a new meta-algorithm, which we call \emph{first-order rejection sampling} (\AlgWSC); see \cref{alg:fors}.
This algorithm aims at simulating rejection sampling using only first-order (gradient) queries.
We give consequences of our method for diffusion sampling and for log-concave sampling.

\paragraph{Diffusion sampling}
We show the following results, where the error is measured in the bounded Lipschitz metric~\eqref{def:BL-metric}, $\epssc$ denotes the $L^2$-error of the score estimates (\cref{def:score-error}), and $\dstar$ is the \emph{intrinsic} dimension of the data distribution (\cref{def:intrinsic}), which is always bounded by the embedding dimension $d$.
\begin{itemize}
    \item Under \textbf{minimal data assumptions}---namely, $\pdata$ has a finite second moment $\mathsf M_2^2$---we obtain $\delta$ error in $O(\dstar\log^3((d+\mathsf M_2^2)/\delta))$ queries with $\Capx=O(1)$.
\end{itemize}
This result strictly improves upon all prior results in the literature in this setting~\citep[e.g.,][]{chen2023improved,Ben+24Diffusion,ConDurSil25KLDiffusion,azangulov2024convergence, li2024adapting, li2025dimension, liang2025low, PotAzaDel25Manifold, tang2025adaptivity, HuaWeiChe26LowDim}.

Beyond this setting, many works in the literature aim to sample with a number of steps which is sublinear in the dimension~\citep{Chen+23FlowODE, jiao2024instance,jiao2025optimal, zhang2025sublinear}.
We also incorporate these advances into our framework. \fcedit{Specifically, we show that under a non-uniform $L$-Lipschitz condition (with respect to the \emph{Frobenius norm}, \cref{asmp:Lip-Frob}--\ref{asmp:Lip-DDPM}), we obtain $\delta$ error in $O(L\log^3((d+\mathsf M_2^2)/\delta))$ steps, with $\Capx = O(1)$. This result is ``almost dimension-free'' and it also recovers our previous result, as the non-uniform $L$-Lipschitz condition always holds with $L\leq \wt{O}(\dstar)$. Further, it also improves upon prior results in the setting with a non-uniform $L$-Lipschitz condition under operator norm (\cref{asmp:non-unif-Lip}), as the implied complexity is $\widetilde O(\min\{d_\star^{2/3}L^{1/3}, \sqrt{dL}\})$.}


\paragraph{Log-concave sampling}
For sampling from log-concave (and isoperimetric) densities with gradient evaluations of the log-density, we recover state-of-the-art results from~\citet{FanYuaChe23ImprovedProx}, except that \emph{we do not require density evaluations}; see \cref{sec:log_concave}.

\subsection{Related work}

As we do not have space to survey the vast literature on diffusion model guarantees, we focus on the ones most relevant to our work.

\paragraph{High-accuracy diffusion sampling}
Prior to this work, \citet{huang2024reverse,wainwright2025score} studied diffusion sampling with additional \emph{zeroth-order} queries. Specifically, given access to estimates of the unnormalized log-densities with bounded error, query complexity bounds of $\tO(d^2\log(1/\delta))$~\citep{huang2024reverse} and $\tO(\sqrt{d}\log^3(1/\delta))$~\citep{wainwright2025score} were shown under additional Lipschitz conditions on the score functions. Both works resort to standard high-accuracy sampling methods (e.g., Metropolis-adjusted Langevin). However, estimating the densities can be extremely challenging in practice. Moreover,~\citet{Hua+25QuantizedDiffusion} propose a method based on learning a quantized score, which requires changing the score matching objective.

\paragraph{Concurrent work} In concurrent work,~\citet{gatmiry2026high} also obtain a high-accuracy guarantee for diffusion sampling.
\fcedit{
  Assuming that $\pdata$ is a Gaussian convolution, i.e., $\pdata = p_\star * \normal{0,\sigma^2 I}$ where $p_\star$ is supported on a ball of radius $R$, they obtain a query complexity of $\wt O((R/\sigma)^2 \log^2(1/\delta))$ assuming that the score errors are bounded in the \emph{sub-exponential} norm.
  Our results also imply a query complexity of $\wt O((R/\sigma)^2 \log^3(1/\delta))$ through the intrinsic dimension (\cref{ssec:intrinsic}) while only requiring $L_2$-score errors.}
We defer a lengthier comparison to
\cref{sec:concurrent}.

\paragraph{Log-concave sampling}
Our application to log-concave sampling is based on the proximal sampler~\citep{LeeSheTia21RGO, Chen+22ProxSampler}, and our results can be compared to~\citet{AltChe24Warm, FanYuaChe23ImprovedProx}.
See the book draft~\citet{Chewi26Book} for an overview of the subject.

\subsection{Notation}


We make use of the following divergences between probability distributions: the \emph{total variation} (TV) distance $\Dtv{\mu}{\nu} \deq \frac{1}{2}\int|\mu-\nu|$; the \emph{Hellinger distance} $\Dhels{\mu}{\nu} \deq \frac12\int (\sqrt\mu-\sqrt\nu)^2$; the \emph{KL divergence} $\Dkl{\mu}{\nu} \deq \En_\mu\log(\mu/\nu)$; the \emph{chi-squared divergence} $\Dchis{\mu}{\nu} \deq \E_\mu (\mu/\nu)^2 - 1$; and the \emph{Wasserstein distance} $W_2^2(\mu,\nu) \deq \inf_{\gamma~\text{coupling of }\mu,\nu} \En_{(X,Y)\sim\gamma}\|X-Y\|^2$.
In addition, we consider the bounded Lipschitz metric which metrizes weak convergence:
\begin{align}
\label{def:BL-metric}
    \Dbl{\mu}{\nu} \deq \sup\{\E_\mu f - \E_\nu f : |f| \le 1,\,\|f\|_{\rm Lip} \le 1\}.
\end{align}
We define
\begin{align}\notag
    \pclip[\B]{x}\deq \max\crl{-B,\min\crl{B,x}}\,, \qquad \forall x\in\RR\,.
\end{align}
We use $\leqsim$ and $O(\cdot)$ to hide absolute constants, i.e., $f\leqsim g$ (and $f=O(g)$) if there is an absolute constant such that $f\leq Cg$.
The notation $\wt O(\cdot)$ hides logarithmic factors.


\section{Background on diffusion models}
\input{sec_prelim.tex}






\section{Key subroutine: Gaussian tilts}\label{sec:gaussian_tilts}
\input{gaussian_tilts}

\section{Diffusion sampling}\label{sec:diffusion}
\input{sec_DM.tex}

\section{Log-concave sampling}\label{sec:log_concave}
\input{sec_log_concave}


\section{Conclusion}


In this work, we have presented high-accuracy algorithms for diffusion sampling which operate under minimal data assumptions (\cref{thm:DM-intrinsic}), with improved dimension dependence under a Lipschitz score assumption (\cref{thm:DM-Lip}).
With regards to the dependence on the target accuracy, our results improve exponentially over all prior works.
Our framework also has implications for log-concave sampling (\cref{sec:log_concave}) using only first-order queries. 

Although this is a primarily theoretical work, we are working toward implementation and experimental evaluation, which will be left for future work.

\icml{
\section{Impact Statement}


Our work provides new algorithms for diffusion sampling with potential speed-ups.
This could lead to many potential impacts, none of which we feel the need to specifically highlight here.
}

\arxiv{
  \section*{Acknowledgments}
  We thank Sam Power for bringing to our attention useful references.
  We acknowledge support from ARO through award W911NF-21-1-0328, Simons Foundation, and the NSF through awards DMS-2031883 and PHY-2019786,  the DARPA AIQ program, and AFOSR FA9550-25-1-0375. CD is supported by a Simons Investigator Award, a Simons Collaboration on Algorithmic Fairness, ONR MURI grant N00014-25-1-2116, and ONR grant N00014-25-1-2296.
}





\bibliography{ref.bib}
\icml{
\bibliographystyle{icml2026}
}

\appendix
\newpage
\icml{
\onecolumn
}


\section{Discussion}
\input{sec_compare.tex}

\section{Additional notation and technical tools}
\input{tools.tex}


\input{appdx_fors}
\section{Proofs for Section~\ref{sec:gaussian_tilts}}\label{appdx:proof-gaussian-tilts}
\input{appdx_RGO.tex}

\section{Structural properties of the diffusion process}\label{appdx:structure}
\input{appdx_structural.tex}
\section{Proofs for Section~\ref{sec:diffusion}}
\input{appdx_DM_path.tex}


\section{Log-concave sampling}\label{appdx:log-concave}
\input{appdx_log_concave}



\end{document}
